---
title: "Converting gene/protein annotation into annother"
author: Scott Lyons Ph.D
output: html_document
---

I put together these few functions because as a person who works mostley with mouse data, I sometimes need to convert to human gene symbol or accession ID's in order to get pathway analysis functions to work that don't natively incorporate mouse pathway terms. I'll include a reference link, but none of this is super novel. I just wanted to make it more straightforward to use. I find myself rummaging through the biomaRt literature everytime i need to convert a mouse gene symbol to a human gene symbol because I can never remember the names of the functions. So here I made a very straightforward method, using condensed functions, to convert from one gene/protein annotation type to another; either in the same species or another.


Load required packages.

```{r echo = TRUE, warning=FALSE}

library(tidyverse)
library(janitor)
library(biomaRt)

```


If you've never used biomaRT it can be downloaded here.

```{r echo = FALSE, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("biomaRt")

```


Read in example data.

```{r echo = TRUE}
df = read.csv('example gene conversion dataset.csv')
```

Add the two databases that you will be using. If you are doing a same species conversion, you still need to add the dataabse name twice.

```{r echo = TRUE}
org.db = c("mmusculus_gene_ensembl", "hsapiens_gene_ensembl")
```


Function that builds the biomaRt databases, given the species specified from org.db.

```{r echo = TRUE}
annotation_databases = function(dataset_lsts){
  
  species1 = useEnsembl("ensembl",  host = "https://uswest.ensembl.org", dataset = org.db[1])
  species2 = useEnsembl("ensembl", host = "https://uswest.ensembl.org", dataset = org.db[2])
  
  lst = list(species1, species2)
}

annotation = annotation_databases(org.db)
```


You can quickly look up the attributes for the annotation databases that you are going to query. 

```{r echo = TRUE}
lst_attriubtes = lapply(annotation, function(x) listAttributes(x))

lapply(lst_attriubtes, head)

```


The function (getLDS) that retrieves the annotation from the 'from' attributes_lst element and convert to the 'to' list element. Make a note that the 'to' element can have multiple attributes. 
I'm going to add some examples of the attribute names and what those look like below. There are so many different types of gene annotation that what I, for example, might call gene name is actually gene symbol. And to me it's a very annoying process that wastes time when I have to look up the exact annotation definition that I'm looking for each time.
I've also found out that the Ensembl biomart server isn't designed to handle large queries. There's apparently a 5 minute time limit (https://support.bioconductor.org/p/9135542/). So if you are getting 'timed out' errors you can either reduce the number of values your converting 'from' by filtering first, or reduce the number of 'to' attributes. 

```{r echo = TRUE}
annotation_convert = function(atribute_lst ,annotation_datasets){
  
  converged = getLDS(attributes = attributes_lst[[1]], filters = attributes_lst[[1]], values = df$gene_symbol, mart = annotation[[1]], 
                     attributesL = attributes_lst[[2]], martL = annotation[[2]], uniqueRows = T)
}

attributes_lst = lst('from' = c('mgi_symbol'),  'to' = c('hgnc_symbol' ,'ensembl_gene_id', 'ensembl_peptide_id'))

convert_annotation = annotation_convert(attributes_lst, annotation)

lapply(convert_annotation, head)

```



```{r}
sessionInfo()
```


